Theoretical aspects in gravity modeling

### 2018-04-17 manuscript/letter.tex
coordinate systems

\title[Gravity disturbance or gravity anomaly?]
      {Gravity disturbance or gravity anomaly?}

@@

The resultant of gravitational force and centrifugal force acting 
on a body at rest on the Earth's surface is called 
\textit{gravity vector} and its intensity is called simply 
\textit{gravity} \citep{heiskanen-moritz1967,
hofmann-wellenhof-moritz2005}.
In the case of gravimetry on moving platforms (e.g., airplanes,
helicopters, marine vessels), there are additional
non-gravitational accelerations due to the vehicle motion, 
such as Coriolis acceleration and high-frequency vibrations 
\citep{glennie-etal2000,nabighian-etal2005-grav,baumann-etal2012}.
Geophysicists use gravity for estimating the Earth's 
internal density distribution whereas geodesists use
gravity to estimate the geoid \citep{li2001}.
Hence, geophysicists are usually interested 
in the gravitational component of the observed gravity, 
which is produced by the Earth's internal 
density distribution.
%In applied geophysics, gravity surveys are usually conducted 
%over local or regional areas on the Earth's surface for the purpose of 
%characterizing geological structures located within the
%crust and upper mantle.
%
%Consequently, geophysicists are only interested in the
%particular gravitational component of gravity which 
%is produced by these target geological structures.
The first step of the procedure for isolating this 
gravitational component consists in removing the non-gravitational 
effects due to the vehicle motion and also the time variations 
such as Earth tides, instrumental drift and barometric 
pressure changes, for example.
If these effects are properly removed, the resultant 
gravity data can be considered as the sum of a 
centrifugal component due to the Earth's rotation and
a gravitational component produced by the whole Earth's
internal density distribution.
The isolation of this particular gravitational component 
and its subsequent use for estimating density 
distributions related to geological structures in subsurface 
are the main goals in applied geophysics 
\citep{blakely1996}.

Based on well-established concepts of the literature,
we present a discussion aiming at bringing some
light to the following question: in geophysical applications,
should we use the gravity disturbance or gravity anomaly?
It seems that this theoretical issue has been 
debated within the scientific community from a 
more geodetic than geophysical point of view
\citep{lafehr1991,chapin1996,li2001,fairhead2003,
hackney-featherstone2003,hinze2005}.
Our reasoning suggests that the gravity 
disturbance is more appropriated than gravity anomalies for 
approximating the gravitational effect produced by the Earth's 
internal density distribution.

@@

Traditionally, the Earth's gravity field is approximated 
by the gravity field produced by a geocentric and rigid ellipsoid 
of revolution, which has the minor axis $b$ 
coincident with the mean rotating axis of the Earth $Z$, the 
same total mass (including the atmosphere) and also the
same angular velocity of the Earth \citep{heiskanen-moritz1967,
vanicek1987,hofmann-wellenhof-moritz2005,torge2012}.
Another characteristic of this model is that its
limiting surface coincides with a particular equipotential 
of its own gravity field.
Here, we follow \citep{torge2012} and call this model as
\textit{normal Earth}.

Similarly to the gravity vector and gravity, 
the resultant of the virtual 
gravitational and centrifugal forces exerted by the normal
Earth on a body at rest at a point $P$ is called 
\textit{normal gravity vector} and its intensity is called 
simply \textit{normal gravity}.
In geodesy, any model used to represent the normal gravity field
can be arbitrarily defined for the only purpose
of keeping the difference from the actual gravity field as small 
as possible \citep{vanicek1987}.

@@

The normal Earth gives rise to the \textit{geodetic coordinate system}
\citep{heiskanen-moritz1967, soler1976, torge2012, bouman_etal2013}.
In this coordinate system, the position of a point $P$
is defined by the \textit{geometric height} $h$, 
\textit{geodetic latitude}
$\varphi$ and \textit{longitude} $\lambda$ (Figure \ref{fig:fig1}).
Geodetic coordinates $(h, \varphi, \lambda)$ can be easily 
converted into geocentric Cartesian coordinates $(X, Y, Z)$
(Figure \ref{fig:fig1}).
The plane containing the point $P$, the axis $Z$ and
the origin $O$ of this \textit{geocentric Cartesian coordinate system}
is called meridian plane (gray plane in Figure \ref{fig:fig1}).
At a given point $(h, \varphi, \lambda)$, there are three 
unit vectors (Figure \ref{fig:fig1}) given by \citep{soler1976}:
\begin{equation}
\begin{split}
\hat{\mitbf{u}} &= 
\begin{bmatrix}
\cos\varphi \, \cos\lambda \\
\cos\varphi \, \sin\lambda \\
\sin\varphi
\end{bmatrix} \\
\hat{\mitbf{v}} &= 
\begin{bmatrix}
-\sin\varphi \, \cos\lambda \\
-\sin\varphi \, \sin\lambda \\
\cos\varphi
\end{bmatrix} \\
\hat{\mitbf{w}} &= 
\begin{bmatrix}
-\sin\lambda \\
\cos\lambda \\
0
\end{bmatrix}
\end{split} \: ,
\label{eq:unit-vectors}
\end{equation}
Notice that these unit vectors are mutually orthogonal and
that the normal gravity vector $\mitbf{\gamma}_{P}$ at a point $P$
is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ at $P$.

\section{Gravity disturbance}

@@

In a local- or regional-gravity study, 
geophysicists commonly use a topocentric Cartesian coordinate system
with origin at a point $P$ and axes $x$, $y$ and $z$ defined
by the unit vectors $\hat{\mitbf{v}}_{P}$,
$\hat{\mitbf{w}}_{P}$ and $-\hat{\mitbf{u}}_{P}$ 
(equation \ref{eq:unit-vectors}), respectively
(Figures \ref{fig:fig1} and \ref{fig:fig2}).
In this coordinate system, the observed gravity vector
$\mitbf{g}_{i}$, at a point $(x_{i}, y_{i}, z_{i})$, 
$i = 1, ..., N$, can be represented by
\begin{equation}
\mitbf{g}_{i} = \mitbf{\gamma}_{i} + \Delta \mitbf{g}_{i} \: ,
\label{eq:gravity-vector}
\end{equation}
where $\mitbf{\gamma}_{i}$ and $\Delta \mitbf{g}_{i}$
are, respectively, the normal gravity vector and a
disturbing gravitational attraction produced by the anomalous 
masses at the point $(x_{i}, y_{i}, z_{i})$.

For each point $(x_{i}, y_{i}, z_{i})$ in the topocentric
Cartesian coordinate system (Figure \ref{fig:fig2}b), there is
a corresponding point $(h_{i}, \varphi_{i}, \lambda_{i})$
in the geodetic coordinate system (Figure \ref{fig:fig1}).
Here, we opted for omitting the equations to convert a point
$(x_{i}, y_{i}, z_{i})$ into a point $(h_{i}, \varphi_{i}, \lambda_{i})$
and vice versa. These equations, however, may be easily
found in the literature \citep[e.g.,][]{heiskanen-moritz1967, soler1976, torge2012, bouman_etal2013}.

@@

\begin{figure}
%    \vspace{5.5cm}
    \includegraphics{figures/cartesian-geodetic-systems.png}
    \caption{Schematic representation of the geodetic coordinate
    system defined by an oblate ellipsoid with semi-minor axis $b$, 
    coincident with the mean Earth's rotation axis, and a semi-major
    axis $a$. In this coordinate system, the position of a point is
    determined by the geometric height $h$, the geodetic latitude 
    $\varphi$ and longitude $\lambda$. 
    The Earth's center of mass is represented 
    by $O$, $P$ represents a point $(h, \varphi, \lambda)$ and 
    $P^{\prime}$ represents the projection of $P$ on the plane $XY$ 
    (Equatorial plane). The plane containing $O$, $P$ and 
    $P^{\prime}$ is represented by the gray triangle in (a) and (b). 
    The unit vectors 
    $\hat{\mitbf{u}}_{P}$, $\hat{\mitbf{v}}_{P}$ and 
    $\hat{\mitbf{w}}_{P}$ define three mutually orthogonal 
    directions at each point $P$.
    In (b), the point $Q$ represents the projection of $P$ on the 
    ellipsoid surface, at the same latitude $\varphi$ and 
    longitude $\lambda$.}
  \label{fig:fig1}
\end{figure}

\begin{figure}
    \includegraphics{figures/local-system.png}
    \caption{(a) Schematic representation of the gravity vector
    $\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$,
    disturbing gravitational attraction $\Delta \mitbf{g}_{P}$
    (equation \ref{eq:gravity-vector}) and the unit vector 
    $\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) 
    at the point $P$ and also the normal gravity vector
    $\mitbf{\gamma}_{Q}$ at the point $Q$ on the reference ellipsoid.
    At the point $P$ the normal gravity vector $\mitbf{\gamma}_{P}$ 
    is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ and the 
    normal gravity $\mitbf{\gamma}_{Q}$ at the point $Q$ is opposite
    the another unit vector $\hat{\mitbf{u}}_{Q}$, which is
    close to $\hat{\mitbf{u}}_{P}$.
    (b) Schematic representation of a topocentric Cartesian 
    coordinate system with origin at a point $P$ and $x$, $y$ and $z$ 
    axes defined by the unit vectors $\hat{\mitbf{v}}_{P}$, 
    $\hat{\mitbf{w}}_{P}$ and $-\hat{\mitbf{u}}_{P}$ 
    (Figure \ref{fig:fig1}), respectively.}
  \label{fig:fig2}
\end{figure}

>>>>

\title[Theoretical aspects in gravity modeling]
      {Theoretical aspects in gravity modeling}

@@

The resultant of gravitational force and centrifugal force acting 
on a body at rest on the Earth's surface is called 
\textit{gravity vector} and its intensity is called 
\textit{gravity} \citep{heiskanen-moritz1967,
hofmann-wellenhof-moritz2005}.
In the case of gravimetry on moving platforms (e.g., airplanes,
helicopters, marine vessels), there are additional
non-gravitational accelerations due to the vehicle motion, 
such as Coriolis acceleration and high-frequency vibrations 
\citep{glennie-etal2000,nabighian-etal2005-grav,baumann-etal2012}.
Geophysicists use gravity for estimating the Earth's 
internal density distribution whereas geodesists use
gravity to estimate the geoid \citep{li2001}.
Hence, geophysicists are usually interested 
in the gravitational component of the observed gravity, 
which is produced by the Earth's internal 
density distribution.

The first step of the procedure for isolating this 
gravitational component consists in removing the non-gravitational 
effects due to the vehicle motion and also the time variations 
such as Earth tides, instrumental drift and barometric 
pressure changes, for example.
If these effects are properly removed, the resultant 
gravity data can be considered as the sum of a 
centrifugal component due to the Earth's rotation and
a gravitational component produced by the whole Earth's
internal density distribution.
The isolation of this particular gravitational component 
and its subsequent use for estimating density 
distributions related to geological structures in subsurface 
are the main goals in applied geophysics \citep{blakely1996}.
The computation of the gravitational effect produced by
the geological structures is commonly known in geophysics as 
\textit{gravity modeling}.
This term is different from \textit{gravity field modeling}, which
is commonly used in geodesy to denote the characterization of the 
gravity field in a local, regional or global scale.

Based on well-established concepts of the literature,
we present a discussion aiming at bringing some
light to the following question: in geophysical applications,
should we use the gravity disturbance or gravity anomaly?
It seems that this theoretical issue has been 
debated within the scientific community from a 
more geodetic than geophysical point of view
\citep{lafehr1991,chapin1996,li2001,fairhead2003,
hackney-featherstone2003,hinze2005}.
Our reasoning suggests that the gravity 
disturbance is more appropriated than the gravity anomaly 
for geophysical purposes.

@@

Traditionally, the Earth's gravity field is approximated 
by the gravity field produced by a reference ellipsoid
(or level ellipsoid), which is rigid and geocentric.
This reference ellipsoid has the minor axis $b$ 
coincident with the mean rotating axis of the Earth $Z$, the 
same total mass (including the atmosphere) and also the
same angular velocity of the Earth \citep{heiskanen-moritz1967,
vanicek1987,hofmann-wellenhof-moritz2005,torge2012}.
Another characteristic of this model is that its
limiting surface coincides with a particular equipotential 
of its own gravity field.
Here, we follow \citep{torge2012} and call this model as
\textit{normal Earth}.

Similarly to the gravity vector and gravity, 
the resultant of the virtual 
gravitational and centrifugal forces exerted by the normal
Earth on a body at rest at a point $P$ is called 
\textit{normal gravity vector} and its intensity is called 
\textit{normal gravity}.
In geodesy, any model used to represent the normal gravity field
can be arbitrarily defined for the only purpose
of keeping the difference from the actual gravity field as small 
as possible \citep{vanicek1987}.

@@

\section{Terrestrial Reference Systems used in gravity modeling}

For geophysical purposes, there are three important Terrestrial Reference Systems used in gravity modeling. 
They rotate with the Earth and are used for describing
positions and movements of objects on and close to the Earth’s surface
\citep{torge2012}.

The first one is a geocentric system of Cartesian coordinates having 
the $Z$-axis coincident with the mean Earth's rotational axis,
the $X$-axis pointing to the Greenwich meridian and the $Y$-axis
directed so as to obtain a right-handed system (Figure \ref{fig:fig1}).
This reference system can be found in the literature with different names: Mean Terrestrial System \citep[e.g.,][]{soler1976}, Earth-fixed geocentric Cartesian system \citep[e.g.,][]{torge2012} 
or Earth-centered Earth-fixed system \citep[e.g.,][]{bouman_etal2013}, for example. Here, we opted for simply using the term Geocentric Cartesian System (GCS).

Other important reference system is a geocentric system of 
geodetic coordinates, which is defined by the reference ellipsoid 
used in the Normal Earth model \citep{heiskanen-moritz1967, soler1976, 
torge2012, bouman_etal2013}. 
In this coordinate system, the position of a point 
is defined by the \textit{geometric height} $h$, 
\textit{geodetic latitude} $\varphi$ and \textit{longitude} $\lambda$ (Figure \ref{fig:fig1}).
For convenience, we call this system Geocentric Geodetic System
(GGS). 
At a given point $(h, \varphi, \lambda)$, there are three 
mutually-orthogonal unit vectors (Figure \ref{fig:fig1}) given by \citep{soler1976}:
\begin{equation}
\begin{split}
\hat{\mitbf{u}} &= 
\begin{bmatrix}
\cos\varphi \, \cos\lambda \\
\cos\varphi \, \sin\lambda \\
\sin\varphi
\end{bmatrix} \\
\hat{\mitbf{v}} &= 
\begin{bmatrix}
-\sin\varphi \, \cos\lambda \\
-\sin\varphi \, \sin\lambda \\
\cos\varphi
\end{bmatrix} \\
\hat{\mitbf{w}} &= 
\begin{bmatrix}
-\sin\lambda \\
\cos\lambda \\
0
\end{bmatrix}
\end{split} \: .
\label{eq:unit-vectors}
\end{equation}
The required equations to convert coordinates $(h, \varphi, 
\lambda)$ referred to the GGS into coordinates $(X, Y, Z)$ referred to 
the GCS (Figure \ref{fig:fig1}) and vice versa can be easily found in 
the literature \citep[e.g.,][]{heiskanen-moritz1967, torge2012, 
bouman_etal2013}.

PAREI AQUI

In a local or regional study, geophysicists commonly use a topocentric 
Cartesian coordinate system (TCS) with origin at a point $P$ on or 
close to the Earth's surface and axes $x$, $y$ and $z$ (Figure 
\ref{fig:fig2}). In the TCS, the axes $x$ and $y$ are parallel to the unit vectors $\hat{\mitbf{v}}_{P}$ $\hat{\mitbf{w}}_{P}$, respectively. 
The $z$-axis, however, is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) and points downward.

\section{Basic concepts}

@@

In this coordinate system, the observed gravity vector
$\mitbf{g}_{i}$, at a point $(x_{i}, y_{i}, z_{i})$, 
$i = 1, ..., N$, can be represented by
\begin{equation}
\mitbf{g}_{i} = \mitbf{\gamma}_{i} + \Delta \mitbf{g}_{i} \: ,
\label{eq:gravity-vector}
\end{equation}
where $\mitbf{\gamma}_{i}$ and $\Delta \mitbf{g}_{i}$
are, respectively, the normal gravity vector and a
disturbing gravitational attraction produced by the anomalous 
masses at the point $(x_{i}, y_{i}, z_{i})$.

For each point $(x_{i}, y_{i}, z_{i})$ in the topocentric
Cartesian coordinate system (Figure \ref{fig:fig2}b), there is
a corresponding point $(h_{i}, \varphi_{i}, \lambda_{i})$
in the geodetic coordinate system (Figure \ref{fig:fig1}).

@@

\begin{figure}
%    \vspace{5.5cm}
    \includegraphics{figures/cartesian-geodetic-systems.png}
    \caption{Schematic representation of the Geocentric Cartesian System (GCS) and the Geocentric Geodetic System (GGS). The GCS has the $Z$-axis coincident with the mean Earth's rotational axis,
    the $X$-axis pointing to the Greenwich meridian and the $Y$-axis
    directed so as to obtain a right-handed system. The GGS is      defined by an oblate ellipsoid with semi-minor axis $b$, 
    coincident with the $Z$-axis of GCS, and a semi-major
    axis $a$. In this coordinate system, the position of a point is
    determined by the geometric height $h$, the geodetic latitude 
    $\varphi$ and longitude $\lambda$. 
    The Earth's center of mass is represented 
    by $O$, $P$ represents a point $(h_{P}, \varphi_{P}, \lambda_{P})$ and $P^{\prime}$ its projection onto the plane $XY$ 
    (Equatorial plane). The plane containing $O$, $P$ and 
    $P^{\prime}$ is represented in gray in (a) and (b). 
    The unit vectors $\hat{\mitbf{u}}_{P}$, $\hat{\mitbf{v}}_{P}$ and 
    $\hat{\mitbf{w}}_{P}$ define three mutually orthogonal 
    directions at $P$ (equation \ref{eq:unit-vectors}).
    In (b), $Q$ represents a point $(h_{Q}, \varphi_{Q}, \lambda_{Q})$,
    which is the projection of $P$ onto the reference ellipsoid, at
    the same latitude and longitude ($\varphi_{Q} = \varphi_{P}$ and $\lambda_{Q} = \lambda_{P}$).}
  \label{fig:fig1}
\end{figure}

\begin{figure}
    \includegraphics{figures/local-system.png}
    \caption{(a) Schematic representation of the gravity vector
    $\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$,
    disturbing gravitational attraction $\Delta \mitbf{g}_{P}$
    (equation \ref{eq:gravity-vector}), unit vector 
    $\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) and
    the normal gravity vector $\mitbf{\gamma}_{Q}$ at the point $Q$ on the reference ellipsoid.
    At $P$, the normal gravity vector $\mitbf{\gamma}_{P}$ 
    is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ and, at $Q$, the normal gravity $\mitbf{\gamma}_{Q}$ is opposite
    to the unit vector $\hat{\mitbf{u}}_{Q}$. These unit vectors are
    close to each other.
    (b) Schematic representation of a topocentric Cartesian 
    coordinate system with origin at a point $P$. The axes $x$
    and $y$ are parallel to the unit vectors $\hat{\mitbf{v}}_{P}$
    and $\hat{\mitbf{w}}_{P}$ (equation \ref{eq:unit-vectors} and
    Figure \ref{fig:fig1}), respectively. The $z$ axis is opposite
    to the unit vector $\hat{\mitbf{u}}_{P}$ (equation 
    \ref{eq:unit-vectors} and Figure \ref{fig:fig1}) and points downward.}
  \label{fig:fig2}
\end{figure}
